---
output: github_document
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.align='center'
)
```

# CEA

R package for contrastive eigen analysis.

<!-- badges: start -->
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

### Authors

Yunhui Qi, Peng Liu, Yumou Qiu

### Contact

qyh@iastate.edu (Yunhui Qi)

## Installation

You can install the development version of CEA from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("YunhuiQi/CEA")
```

## Website

Check the vignette at https://yunhuiqi.github.io/CEA/.


## Introduction

The goal of CEA is to conduct contrastive eigen analysis including contrastive principal component (cPCA) and its sparse version (scPCA), contrastive cross-covariance analysis (3CA) and its sparse version (3CA). (s)cPCA can be implemented by function `cPCA`. (s)3CA can be implemented by function `cCCA`.


```{r load libraries for this intro, message=FALSE}
library(CEA)
library(ggpubr)
library(ggplot2)
library(ggsci)
```


### cPCA 

`eta_tuning()`

* Input

  + `Xt` data frame for treatment group, samples in rows, variables in columns. If PCA is applied to correlation matrix, Xt should be scaled and centered.  
  + `Xc` data frame for control group, samples in rows, variables in columns. If cPCA is applied to correlation matrix, Xc should also be scaled and centered.  
  + `eta_range` a vector containing candidate eta range to be tuning from.  
  + `plot` a logical quantity indicating whether plot the ARI(nuisance) and silhouette score.  
  + `sparse_trt` parameter controlling the sparsity of treatment sparse PCA, if NULL, use cPCA, else use scPCA. For use of scPCA, if vector, refer sumabss in PMD.cv from package PMA, if single number, refer sumabs in PMD in package PMA.  
  + `sparse_ctst` parameter controlling the sparsity of scPCA, if NULL, use cPCA, else use scPCA. For use of scPCA, if vector, refer sumabss in PMD.cv from package PMA, if single number, refer sumabs in PMD in package PMA.  
  + `km_cluster` number showing number of nuisance clusters defined from kmeans clustering on treatment PCA. 
  + `ckm_cluster` number of interesting clusters defined from kmeans clustering on cPCA/scPCA. 


* Output

  + A list containing selected value of $\eta$  `eta_opt`, the vector of ARI(nuisance) `nuisance_ARI` and the vector of Silhouette `silhouette`.


`cPCA()`

* Input

  + `Xt` data frame for treatment group, samples in rows, variables in columns.
  + `Xc` data frame for control group, samples in rows, variables in columns.
  + `eta` tuning parameter controlling how much control variation should be contrasted off from the treatment variation. It can be theo `eta_opt` from function eta_tuning(). 
  + `sparse_ctst` parameter controlling the sparsity of scPCA, if NULL, use cPCA, else use scPCA. For use of scPCA, if vector, refer sumabss in PMD.cv from package PMA, if single number, refer sumabs in PMD in package PMA.  
  + `sparse_trt` parameter controlling the sparsity of treatment sparse PCA, if NULL, use cPCA, else use scPCA. For use of scPCA, if vector, refer sumabss in PMD.cv from package PMA, if single number, refer sumabs in PMD in package PMA.  
  + `sparse_ctrl` parameter controlling the sparsity of control sparse PCA, if NULL, use cPCA, else use scPCA. For use of scPCA, if vector, refer sumabss in PMD.cv from package PMA, if single number, refer sumabs in PMD in package PMA.


* Output

  + A list containing contrastive scores `cPC` and loadings `cU`, treatment scores `PCt` and loadings `Ut`, and control scores `PCc` and loadings `Uc`.

1. We load the simulated data from the `data(intro)`, and use `eta_tuning()` to get optimal value of tuning parameter $\eta$. 


```{r eta tuning for cPCA}
eta_range <- seq(0.1, 5, 0.05)
eta_opt <- eta_tuning(Xt = intro$cPCA$Xt, Xc = intro$cPCA$Xc, eta_range = eta_range, plot = TRUE, sparse_trt = NULL, sparse_ctst = NULL)$eta_opt
```

2. We then input the data and the resulting $\eta$ value to cPCA.

```{r cPCA usage}
# speccify sparsity parameters to be NULL for cPCA
cpca_res <- cPCA(Xt = intro$cPCA$Xt, Xc = intro$cPCA$Xc, eta = eta_opt, sparse_ctst = NULL, sparse_trt = NULL, sparse_ctrl = NULL)
```

3. We get the score plot of cPCA. In comparison, we also get the score plot of treatment PCA. Here, as in the paper, we have nuisance subgroup factor $C$ and interesting subgroup factor $B$. In the score plot, we see the first treatment score is associated with the nuisance subgroups, while the first contrastive score is associated with the interesting subgroups. This validates the removal of nuisance effect and discovery of hidden subgroups by cPCA.

```{r cPCA usage score plot, fig.height=5, fig.width=10}
plot_data <- data.frame(cPC1 = cpca_res$cPC[,1], cPC2 = cpca_res$cPC[,2],
                        PC1 = cpca_res$PCt[,1], PC2 = cpca_res$PCt[,2],
                        intro$cPCA$covariates_t)
cP <- ggplot(plot_data)+
        geom_point(aes(x = cPC1, y = cPC2, color = B, shape = C))+
        labs(title = "Contrastive", color = "Interesting \n subgroups", shape = "Nuisance \n subgroups")+
        theme_bw()+
        theme(panel.grid = element_blank(),
              text = element_text(size = 14),  # Adjust text size
              axis.title = element_text(size = 16),  # Adjust axis title size
              axis.text = element_text(size = 12))
Pt <- ggplot(plot_data)+
        geom_point(aes(x=PC1, y=PC2, color= B, shape = C), show.legend = FALSE)+
        labs(title = "Treatment", color = "Interesting \n subgroups", shape = "Nuisance \n subgroups")+
        theme_bw()+
        theme(panel.grid = element_blank(),
              text = element_text(size = 14),  # Adjust text size
              axis.title = element_text(size = 16),  # Adjust axis title size
              axis.text = element_text(size = 12))
p_score <- ggarrange(Pt,cP, nrow = 1, common.legend = TRUE, legend = "bottom", align = "v")
print(p_score)
```

### scPCA

To illustrate the use of function `cPCA()` for sparse cPCA, we use the barley expression data example. For details of data processing and references, please refer to the paper and `data(barley)`.

1. We load the processed data from the `data(intro)`, and use assigned $\eta = 1$, we also specify the sparsity parameters to be 0.3.

```{r scPCA usage}
set.seed(111)
# specify the sparsity parameters for scPCA
scPCA_res <- cPCA(Xt = intro$scPCA$Xt, Xc = intro$scPCA$Xc, eta = 1, sparse_ctst = 0.3, sparse_trt = 0.3, sparse_ctrl = 0.3)
```


2. We get the score plot of scPCA. In comparison, we also get the score plot of treatment PCA. As described in the paper, Genotype effect is removed by scPCA, and the fungus subgroups is revealed by the first scPCA score.

```{r scPCA usage score plot, fig.height=5, fig.width=10}
PCt <- scPCA_res$PCt
PCc <- scPCA_res$PCc
cPC <- scPCA_res$cPC
Ut <- scPCA_res$Ut
Uc <- scPCA_res$Uc
cU <- scPCA_res$cU

plot_data_t <- data.frame(cPC1 = cPC[,1], cPC2 = cPC[,2],
                          PC1 = PCt[,1], PC2 = PCt[,2],
                          Genotype = intro$scPCA$covariates_t$Genotype,
                          Isolate = intro$scPCA$covariates_t$Isolate,
                          Time_bi = factor(intro$scPCA$covariates_t$Time_bi, levels = c("Early","Late")))
plot_data_c <- data.frame(PC1 = PCc[,1], PC2 = PCc[,2], 
                          Genotype = intro$scPCA$covariates_c$Genotype,
                          Isolate = intro$scPCA$covariates_c$Isolate,
                          Time = intro$scPCA$covariates_c$Time,
                          Time_bi = factor(intro$scPCA$covariates_c$Time_bi, levels = c("Early","Late")))

cP <- ggplot(plot_data_t)+
  geom_point(aes(x = cPC1, y = cPC2, color = Isolate, shape = Genotype), size = 2)+
  theme_bw()+
  labs(title = "Contrastive", color = "Fungus")+
  theme(panel.grid = element_blank(),
        text = element_text(size = 14),  # Adjust text size
        axis.title = element_text(size = 16),  # Adjust axis title size
        axis.text = element_text(size = 12),
        legend.position = "right")+
  scale_color_d3()
Pt <- ggplot(plot_data_t)+
  geom_point(aes(x=PC1, y=PC2, color= Isolate, shape = Genotype ), size = 2)+
  labs(title = "Late", color = "Fungus")+
  theme_bw()+
  theme(panel.grid = element_blank(),
        text = element_text(size = 14),  # Adjust text size
        axis.title = element_text(size = 16),  # Adjust axis title size
        axis.text = element_text(size = 12),
        legend.position = "right")+
  scale_color_d3()
Pc <- ggplot(plot_data_c)+
  geom_point(aes(x=PC1, y=PC2, color= Isolate, shape = Genotype), size = 2)+
  labs(title = "Early", color = "Fungus")+
  theme_bw()+
  theme(panel.grid = element_blank(),
        text = element_text(size = 14),  # Adjust text size
        axis.title = element_text(size = 16),  # Adjust axis title size
        axis.text = element_text(size = 12),
        legend.position = "right")+
  scale_color_d3()

p_score <- ggarrange(Pc, Pt, cP,  nrow=1, ncol=3,  common.legend = TRUE, legend = "bottom")
print(p_score)
```

### 3CA

`cCCA()`

* Input 

  + `Xt` data frame for one data type in treatment group, samples in rows, variables in columns. To conduct analysis based on cross corrlation matrix, each feature should be centered and standardized.    
  + `Xc` data frame for one data type in control group, samples in rows, variables in columns. To conduct analysis based on cross corrlation matrix, each feature should be centered and standardized.  
  + `Yt` data frame for another data type in treatment group, samples in rows, variables in columns. To conduct analysis based on cross corrlation matrix, each feature should be centered and standardized.  
  + `Yc` data frame for another data type in control group, samples in rows, variables in columns. To conduct analysis based on cross corrlation matrix, each feature should be centered and standardized.  
  + `eta` tuning parameter controlling how much control variation should be contrasted off from the treatment variation. If NULL, the estimated eta introduced in paper will be used.
  + `sparse_ctst` parameter controlling the sparsity of s3CA, if NULL, use 3CA, else use s3CA. For use of s3CA, if vector, refer sumabss in PMD.cv from package PMA, if single number, refer sumabs in PMD in package PMA.  
  + `sparse_trt` parameter controlling the sparsity of treatment sparse cross covariance/correlation analysis, if NULL, use cross covariance/correlation analysis, else use sparse cross covariance/correlation analysis. For use of sparse version, if vector, refer sumabss in PMD.cv from package PMA, if single number, refer sumabs in PMD in package PMA.  
  + `sparse_ctrl` parameter controlling the sparsity of control sparse cross covariance/correlation analysis, if NULL, use cross covariance/correlation analysis, else use sparse cross covariance/correlation analysis. For use of sparse version, if vector, refer sumabss in PMD.cv from package PMA, if single number, refer sumabs in PMD in package PMA.  

* Output

  + A list containing contrastive SVD object `CCA_ctst` including singular values `d`, loadings for X `u`, loadings for Y `v`; treatment only SVD object `CCA_trt` and control only SVD object `CCA_ctrl`.



1. We load the simulated data from the `data(intro)`, and get the top estimated latent factor (score) from 3CA. As in the paper, we also simulated the external continuous response from linear model using the target population latent factor with coefficient 1.5. Here we conduct 3CA and get the estimated top latent factor and regress the response on the score to recover the coefficient. 


```{r 3CA usage}
res_3CA <- cCCA(Xc = intro$cCCA$Xc, Yc = intro$cCCA$Yc, 
                Xt = intro$cCCA$Xt, Yt = intro$cCCA$Yt, eta  = NULL,
                sparse_ctst = NULL, sparse_trt = NULL, sparse_ctrl = NULL)
```

2. The coefficient 1.5 is recovered by the estimated score from 3CA, validating that the target latent factor is estimated by 3CA.

```{r 3CA usage regression}
coefficients(lm(intro$cCCA$response ~ intro$cCCA$Xt %*% res_3CA$CCA_ctst$u[,1]))[2]
```


### s3CA

To illustrate the use of function `s3CA()` for sparse 3CA, we use the covid metabolome and proteome data example. For details of data processing and references, please refer to the paper and `data(covid)`.

1. We load the processed data from the `data(intro)`, and use assigned $\eta = 1$, we also specify the sparsity parameters range to use cross validation to select the sparsity parameters. s3CA is implemented to get the score plot. 

```{r s3CA usage}
s3CA_res <- cCCA(Xc = intro$s3CA$Xc, Yc = intro$s3CA$Yc, 
                 Xt = intro$s3CA$Xt, Yt = intro$s3CA$Yt, eta  = 1,
                 sparse_ctst = seq(0.2, 1, len=20), 
                 sparse_trt = seq(0.2, 1, len=40), 
                 sparse_ctrl = seq(0.2, 1, len=40))
```


2. We get the score plot from s3CA. In comparison, we also get the score plot of covid group and non-covid group cross-covariance analysis respectively. As described in the paper, ICU effect is removed by contrasting, and the estimated top latent factor by s3CA is associated with the external continuous variable - platelet counts.

```{r s3CA usage score plot, warning=FALSE, message=FALSE, fig.height=6, fig.width=10}
CCA_ctst <- s3CA_res$CCA_ctst
CCA_ctrl <- s3CA_res$CCA_ctrl
CCA_trt <- s3CA_res$CCA_trt


df_score <- data.frame(cCCX = intro$s3CA$Xt%*%as.matrix(CCA_ctst$u[,1]), 
                       cCCY = intro$s3CA$Yt%*%as.matrix(CCA_ctst$v[,1]),
                       intro$s3CA$covariates_t)
df_score_t <- data.frame(CCtX = intro$s3CA$Xt%*%as.matrix(CCA_trt$u[,1]), 
                         CCtY = intro$s3CA$Yt%*%as.matrix(CCA_trt$v[,1]), 
                         intro$s3CA$covariates_t)

df_score_c <- data.frame(CCcX = intro$s3CA$Xc%*%as.matrix(CCA_ctrl$u[,1]), 
                         CCcY = intro$s3CA$Yc%*%as.matrix(CCA_ctrl$v[,1]), 
                         intro$s3CA$covariates_c)

interested_y <- "Platelet_K.uL"
p_reg <- ggplot(data = data.frame(df_score[,c("cCCY","icu",interested_y)]),
                aes_string(x = "cCCY",y = interested_y))+
  geom_point(aes_string(color = "icu" ))+ 
  geom_smooth(method = "lm", se = TRUE)+
  theme_bw()+
  labs(title = "Contrastive", x = "Yv", color = "")+
  theme(panel.grid = element_blank(),
        text = element_text(size = 14),  # Adjust text size
        axis.title = element_text(size = 16),  # Adjust axis title size
        axis.text = element_text(size = 12),
        legend.position = "right")+
  scale_color_d3()
p_reg_t <- ggplot(data = data.frame(df_score_t[, c("CCtY", "icu",interested_y)]),
                aes_string(x = "CCtY",y = interested_y))+
  geom_point(aes_string(color = "icu" ))+  
  geom_smooth(method = "lm", se = TRUE)+
  theme_bw()+
  labs(title = "COVID", x = "Yv", color = "")+
  theme(panel.grid = element_blank(),
        text = element_text(size = 14),  # Adjust text size
        axis.title = element_text(size = 16),  # Adjust axis title size
        axis.text = element_text(size = 12),
        legend.position = "right")+
  scale_color_d3()
p_reg_c <- ggplot(data = data.frame(df_score_c[, c("CCcY", "icu",interested_y)]),
                aes_string(x = "CCcY",y = interested_y))+
  geom_point(aes_string(color = "icu" ))+ 
  geom_smooth(method = "lm", se = TRUE)+
  theme_bw()+
  labs(title = "Non COVID", x = "Yv", color = "")+
  theme(panel.grid = element_blank(),
        text = element_text(size = 14),  # Adjust text size
        axis.title = element_text(size = 16),  # Adjust axis title size
        axis.text = element_text(size = 12),
        legend.position = "right")+
  scale_color_d3()

p_score <- ggarrange(p_reg_t, p_reg_c, p_reg, common.legend = T, 
                     legend = "bottom", nrow=1, align = "hv")
print(p_score)
```

