---
title: "CEAvignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CEAvignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 10,
  out.width = "100%"
)
```

```{r setup}
library(CEA)
library(ggplot2)
library(ggsci)
library(ggh4x)
library(ggpubr)
library(tidyverse)
library(lme4)
library(lmerTest)
library(ComplexHeatmap)
library(RColorBrewer)
```


# Introduction 

CEA is an R package for contrastive eigen-analysis including contrastive PCA (cPCA) and its sparse version (scPCA), contrastive cross-covariance analysis (3CA) and its sparse version (s3CA).


In this vignette, we will use CEA package to reproduce the analysis in the original paper, including

1. Single dataset eta tuning example, generating Fig.2

2. Simulation for contrastive spectral clustering, generating Table 2.

3. Simulation for sparse contrastive spectral clustering, generating Table 3.

4. Simulation for 3CA, generating Fig.3

5. Simulation for s3CA, generating Fig.S1 and Fig.S2

6. Real data application of s3CA generating Fig.4 and Table S3, Fig S3.


# Simulation

## 1. Single dataset eta tuning example, generating Fig.2

```{r, eta tuning example}
eta_range <-  seq(0, 100, 0.2)
Xt <- scale(sim_data_CSC$Xt, scale = FALSE)
Xa <- scale(sim_data_CSC$Xa, scale = FALSE)
meta_t <- sim_data_CSC$covariates_t
eta_tuning_res <-  eta_tuning_general(Xt = Xt, Xa = Xa, eta_range = eta_range)
eta_opt <- eta_tuning_res$eta_opt
p_tuning <- eta_tuning_res$plot+scale_color_cosmic()+labs(title = expression("ARI/Silhouette vs candidate"~eta))


meta_all <-  setNames(data.frame(matrix(ncol = ncol(meta_t)+3, nrow = 0)), c(colnames(meta_t), "cPC1","cPC2","eta"))
for (eta_use in c(eta_opt, seq(from = min(eta_range), to = max(eta_range), length.out = 3))) {
  meta_tmp <- meta_t
  meta_tmp$cPC1 <-  Xt %*% eigen(cov(Xt)-eta_use*cov(Xa))$vectors[,1]
  meta_tmp$cPC2 <-  Xt %*% eigen(cov(Xt)-eta_use*cov(Xa))$vectors[,2]
  meta_tmp$eta <- paste0("eta==",eta_use)
  meta_all <- rbind(meta_all, meta_tmp)
}
meta_all$eta <- factor(meta_all$eta, levels=c("eta==0","eta==1.6","eta==50","eta==100"))
p_eta <- ggplot(meta_all)+
            geom_point(aes(x = cPC1, y = cPC2, color = interesting, shape = nuisance))+
            facet_wrap(vars(eta), labeller = label_parsed, scales= "free")+
            theme_bw()+
            scale_color_aaas()+
            scale_shape_manual(values = c(1, 15))+
            theme(
              strip.background = element_rect(fill = "lightblue", color = "black"),
              strip.text = element_text(color = "black", face = "bold", size = 14),
              text = element_text(size = 14),  # Adjust text size
              axis.title = element_text(size = 16),  # Adjust axis title size
              axis.text = element_text(size = 12),   # Adjust axis text size
              legend.position = "bottom")
print(p_eta)
# ggsave("../inst/SavedFigures/sim_etaTuningEg.pdf", 
#        plot = ggarrange(p_tuning, p_eta, ncol = 2, widths = c(1,1.6), labels = "AUTO"), 
#        width = 12, height = 6, units = "in")
```

## 2. Simulation for contrastive spectral clustering, generating Table 2 
```{r  simulation cPCA, eval=FALSE}
set.seed(111)
mu2_candidates <- c(1, 2)
p1_candidates <- c(5, 10)
eta_range <-  seq(0.1, 50, 0.2)
Sigmat_candidates <- c("Identity", "Toep(0.4)", "Toep(0.5)")
results <- setNames(data.frame(matrix(ncol = 8, nrow = 0)),
                     c("ARI_ctst_interesting", "ARI_ctst_nuisance",
                       "ARI_trt_interesting", "ARI_trt_nuisance","eta",
                        "mu2", "p1", "Sigmat"))
for (mu2 in mu2_candidates) {
  for (p1 in p1_candidates) {
    for(Sigmat in Sigmat_candidates){
        p2 <- 10
        p <- p1+p2
        Sigma_t <- case_when(Sigmat == "Identity" ~ diag(1, p),
                             Sigmat == "Toep(0.4)" ~ toeplitz(0.4^(0:(p-1))),
                             Sigmat == "Toep(0.5)" ~ toeplitz(0.5^(0:(p-1))))
        res <- sim_CSC(n11 = 50, n12 = 25, n21 = 55, n22 = 30, m1tilde = 20, m2tilde = 30,
                       mu1 = 5, mu2 = mu2, theta1 = -5, theta2 = 6, p1 = p1, p2 = p2,
                       sigma_t = 1, sigma_a = 1,
                       Sigma_t = Sigma_t, Sigma_a = NULL,
                       eta = eta_range, repetition = 50, save_plot = FALSE)
        res$mu2 <- mu2
        res$mu <- abs(mu2-5)
        res$p1 <- p1
        res$Sigmat <- Sigmat
        results <- rbind(results, res)
      }
    }
  }
# save(results, file = "../inst/SavedData/sim_cPCA.RData")
```

```{r  simulation cPCA visualization, eval=TRUE}
load(file = "../inst/SavedData/sim_cPCA.RData")
plot_results <- pivot_longer(results, 1:4, names_to = "Type", values_to = "ARI")
plot_results$Analysis <- sapply(plot_results$Type, function(x){strsplit(x,"_")[[1]][2]})
plot_results$Cluster <- sapply(plot_results$Type, function(x){strsplit(x,"_")[[1]][3]})
plot_results$Analysis <- ifelse(plot_results$Analysis=="ctst","Contrastive","Target-only")
plot_results$Cluster <- ifelse(plot_results$Cluster=="interesting","Interesting","Nuisance")
plot_results$eta <- paste("eta==", plot_results$eta, sep = "")
plot_results$mu <- paste("mu==", plot_results$mu, sep = "")
plot_results$p1 <- paste("p[1]==", plot_results$p1, sep = "")
plot_results <- plot_results %>% group_by(mu, p1, Sigmat, Analysis, Cluster) %>%
  summarise(Average_ARI = mean(ARI), SE_ARI = sd(ARI), covarage_high = mean(ARI >= 0.9), covarage_low = mean(ARI <= 0.1))
print(knitr::kable(plot_results, format = "markdown", digits = 3))
p_cPCA <- ggplot(plot_results, aes(x = Cluster,  y = Average_ARI))+
          geom_point(aes(color = Analysis), stat="identity", position ="dodge")+
          geom_errorbar(aes(ymin = Average_ARI - SE_ARI, ymax = Average_ARI + SE_ARI), linewidth = 0.5) +
          facet_nested(Sigmat ~ mu + p1, labeller = label_parsed)+
          labs(x = "", y="Average ARI")+
          theme_bw()+
          theme(
            panel.grid = element_blank(),
            strip.background = element_rect(fill = "lightblue", color = "black"),
            strip.text = element_text(color = "black", face = "bold", size = 14),
            text = element_text(size = 14),  # Adjust text size
            axis.title = element_text(size = 16),  # Adjust axis title size
            axis.text = element_text(size = 12),   # Adjust axis text size
            legend.position = "bottom"
         )+
         scale_color_aaas()
print(p_cPCA)
# ggsave("../inst/SavedFigures/sim_cPCA.pdf", p_cPCA, width = 6, height = 9, units = "in")
```

## 3. Simulation for sparse contrastive spectral clustering, generating Table 3

```{r simulation scPCA, eval = FALSE}
set.seed(111)
mu2_candidates <- c(1, 2)
p3_candidates <- c(100, 300)
eta_range <-  seq(0.1, 10, 0.5)
Sigmat_candidates <- c("Identity", "Toep(0.4)", "Toep(0.5)")
results <- setNames(data.frame(matrix(ncol = 8, nrow = 0)),
                     c("ARI_ctst_interesting", "ARI_ctst_nuisance",
                     "ARI_trt_interesting", "ARI_trt_nuisance","eta",
                      "mu2", "p3", "Sigmat"))
for (mu2 in mu2_candidates) {
  for (p3 in p3_candidates) {
    for(Sigmat in Sigmat_candidates){
      p1 <- 5
      p2 <- 10
      p <- p1+p2+p3
      Sigma_t <- case_when(Sigmat == "Identity" ~ diag(1, p),
                           Sigmat == "Toep(0.4)" ~ toeplitz(0.4^(0:(p-1))),
                           Sigmat == "Toep(0.5)" ~ toeplitz(0.5^(0:(p-1))))
      res <- sim_sCSC(n11 = 50, n12 = 25, n21 = 55, n22 = 30, m1tilde = 20, m2tilde = 30,
                      mu1 = 5, mu2 = mu2, theta1 = -5, theta2 = 6, p1 = p1, p2 = p2, p3 = p3,
                      sigma_t = 1, sigma_a = 1, Sigma_t = Sigma_t, Sigma_a = NULL,
                      eta = eta_range, repetition = 50)
       res$mu2 <- mu2
       res$mu <- abs(mu2-5)
       res$p3 <- p3
       res$Sigmat <- Sigmat
       results <- rbind(results, res)
     }
   }
}
# save(results, file = "../inst/SavedData/sim_scPCA.RData")
```

```{r  simlation scPCA visualization, eval=TRUE}
load(file = "../inst/SavedData/sim_scPCA.RData")
plot_results <- pivot_longer(results, 1:4, names_to = "Type", values_to = "ARI")
plot_results$Analysis <- sapply(plot_results$Type, function(x){strsplit(x,"_")[[1]][2]})
plot_results$Cluster <- sapply(plot_results$Type, function(x){strsplit(x,"_")[[1]][3]})
plot_results$Analysis <- ifelse(plot_results$Analysis=="ctst","Contrastive","Target-only")
plot_results$Cluster <- ifelse(plot_results$Cluster=="interesting","Interesting","Nuisance")
plot_results$eta <- paste("eta==", plot_results$eta, sep = "")
plot_results$mu <- paste("mu==", plot_results$mu, sep = "")
plot_results$p3 <- paste("p[3]==", plot_results$p3, sep = "")
plot_results <- plot_results %>% group_by(mu, p3, Sigmat, Analysis, Cluster) %>%
  summarise(Average_ARI = mean(ARI), SE_ARI = sd(ARI), covarage_high = mean(ARI >= 0.9), covarage_low = mean(ARI <= 0.1))
print(knitr::kable(plot_results, format = "markdown", digits = 3))
p_scPCA <- ggplot(plot_results, aes(x = Cluster,  y = Average_ARI))+
          geom_point(aes(color = Analysis), stat="identity", position ="dodge")+
          geom_errorbar(aes(ymin = Average_ARI - SE_ARI, ymax = Average_ARI + SE_ARI), linewidth = 0.5) +
          facet_nested(Sigmat ~ mu + p3, labeller = label_parsed)+
          labs(x = "", y="Average ARI")+
          theme_bw()+
          theme(
            panel.grid = element_blank(),
            strip.background = element_rect(fill = "lightblue", color = "black"),
            strip.text = element_text(color = "black", face = "bold", size = 14),
            text = element_text(size = 14),  # Adjust text size
            axis.title = element_text(size = 16),  # Adjust axis title size
            axis.text = element_text(size = 12),   # Adjust axis text size
            legend.position = "bottom"
         )+
         scale_color_aaas()
print(p_scPCA)
# ggsave("../inst/SavedFigures/sim_scPCA.pdf", p_scPCA, width = 6, height = 9, units = "in")
```

## 4. Simulation for 3CA, generating Fig.3

```{r simulation for 3CA, eval=FALSE}
n_candidates <- c(1000, 2000, 3000)
AR_candidates <- c(0.5, 0.7, 0.9)
results <- setNames(data.frame(matrix(ncol = 6, nrow = 0)), 
                    c("est_trt_cca", "est_ctst_cca", "est_ctst_pca", "eta_est", 
                      "SampleSize", "AR"))
for (n in n_candidates) {
  for (AR in AR_candidates){
    res <- sim_3CA(nt = n, na = n, p = 40, q = 40, s = 10, AR = AR,
                   lambda_t = c(6, 4, 2, 0), lambda_a = c(2, 0.2), eta = NULL,
                   lm_sigma = 1, lm_beta = 1.5, repetition = 50)
    rownames(res) <- NULL
    res$SampleSize <- n
    res$AR <- AR
    results <- rbind(results, res)
  }
}
# save(results, file = "../inst/SavedData/sim_3CA.RData")
```

```{r simulation for 3CA visualization}
load("../inst/SavedData/sim_3CA.RData")
plot_results <- pivot_longer(results, 1:3, names_to = "Type", values_to = "Estimation")
plot_results$Analysis <- sub(".*?_", "", plot_results$Type)
plot_results$SampleSize <- paste("n=", plot_results$SampleSize, sep = "")
plot_results$AR <- paste("AR=", plot_results$AR, sep = "")
plot_results$Analysis <- factor(plot_results$Analysis, labels = c("ContrastiveCCA", "ContrastivePCA", "TreatmentCCA"))
## boxplot
p_3CA <- ggplot(data = plot_results)+
            geom_boxplot(aes(x = factor(Analysis, levels=c("ContrastiveCCA", "ContrastivePCA", "TreatmentCCA"), labels=c("3CA", "cPCA", "trtCCA")), y = abs(Estimation), fill = Analysis), show.legend = FALSE)+
            facet_grid(cols = vars(SampleSize), rows = vars(AR))+
            geom_hline(yintercept = 1.5, linetype = "dashed", color="red")+
            scale_y_continuous(breaks = c( 0, 1.5))+
            theme_bw()+
            theme(
              panel.grid = element_blank(),
              strip.background = element_rect(fill = "lightblue", color = "black"),
              strip.text = element_text(color = "black", face = "bold", size = 14),
              text = element_text(size = 14),  # Adjust text size
              axis.title = element_text(size = 16),  # Adjust axis title size
              axis.text = element_text(size = 12),   # Adjust axis text size
              # axis.text.x = element_text(angle = 45, hjust = 1)
            )+
            scale_fill_d3()+
            labs(y = "Estimation of the coefficient in (6.2)",x="Analysis")
print(p_3CA)
# ggsave("../inst/SavedFigures/sim_3CA.pdf", plot = p_3CA, width = 10, height = 10, units = "in")

```


## 5. Simulation for s3CA, generating Fig.S1 and Fig.S2

```{r simulation for s3CA, eval=FALSE}
n_candidates <- c(3000)
AR_candidates <- c(0.5, 0.7, 0.9)
p_candidates <- c(300, 350, 400)
s_candidates <- c(5,10)
results <- setNames(data.frame(matrix(ncol = 7, nrow = 0)), 
                    c("est_trt_cca", "est_ctst_cca", "est_ctst_pca", "eta_est",
                      "Dimension", "AR","Sparsity"))
for (n in n_candidates) {
  for (AR in AR_candidates){
    for (p in p_candidates) {
      for (s in s_candidates) {
        res <- sim_s3CA(nt = n, na = n, p = p, q = p, s = s, 
                        lambda_t = c(6, 4, 2, 0), lambda_a = c(2, 0.2), eta = NULL,
                        AR = AR, lm_sigma = 1, lm_beta = 1.5, repetition = 50)
        rownames(res) <- NULL
        res$Dimension <- p
        res$AR <- AR
        res$Sparsity <- s
        results <- rbind(results, res)
      }
    }
  }
}
# save(results, file = "../inst/SavedData/sim_s3CA.RData")
```

```{r simulation for s3CA visualization}
load("../inst/SavedData/sim_s3CA.RData")
plot_results <- pivot_longer(results, 1:3, names_to = "Type", values_to = "Estimation")
plot_results$Analysis <- sub(".*?_", "", plot_results$Type)
plot_results$Dimension <- paste("p=", plot_results$Dimension, sep = "")
plot_results$AR <- paste("AR=", plot_results$AR, sep = "")
plot_results$Analysis <- factor(plot_results$Analysis, labels = c("ContrastiveCCA", "ContrastivePCA", "TargetCCA"))
plot_results$Sparsity <- paste("# of non-zero elements =", plot_results$Sparsity)

p_s3CA <- ggplot(data = plot_results[plot_results$Sparsity == "# of non-zero elements = 10",])+
      geom_boxplot(aes(x = factor(Analysis, levels=c("ContrastiveCCA", "ContrastivePCA", "TargetCCA"), labels=c("s3CA", "scPCA", "trtsCCA")), y = abs(Estimation), fill = Analysis), show.legend = FALSE)+
      facet_nested(AR~Sparsity+Dimension)+
      geom_hline(yintercept = 1.5, linetype = "dashed", color="red")+
      scale_y_continuous(breaks = c( 0, 1.5))+
      theme_bw()+
      theme(
        panel.grid = element_blank(),
        strip.background = element_rect(fill = "lightblue", color = "black"),
        strip.text = element_text(color = "black", face = "bold", size = 14),
        text = element_text(size = 14),  # Adjust text size
        axis.title = element_text(size = 16),  # Adjust axis title size
        axis.text = element_text(size = 12),   # Adjust axis text size
      )+
      scale_fill_d3()+
      labs(y = "Estimation of the coefficient in (6.2)",x="Analysis")
print(p_s3CA)
# ggsave("../inst/SavedFigures/sim_s3CA_s10.pdf", plot = p_s3CA, width = 10, height = 10, units = "in")
# ggsave("../inst/SavedFigures/sim_s3CA_s5.pdf", plot = p_s3CA, width = 10, height = 10, units = "in")
```


# Real data application of s3CA on COVID19 data.

```{r covid data s3CA}
data(covid)
covariates <- covid$covariates
mtb <- covid$mtb
protein <- covid$protein

# transformation
mtb_log <- log(mtb + 1)
protein_log <- log(protein + 1)

#### 1. Choose correlation or covariance ####
# Use correlation matrix or covariance matrix
cormat <- TRUE
# Use assigned eta and zta
eta_assign <- 1 # or NULL to use eta estimation
#### 2. Choose conditions ####
# Take Covid as conditions 
Xa <- scale(mtb_log[covariates$covid == "Non-Covid", ], scale = cormat)
Ya <- scale(protein_log[covariates$covid == "Non-Covid",], scale = cormat)
Xt <- scale(mtb_log[covariates$covid != "Non-Covid",], scale = cormat)
Yt <- scale(protein_log[covariates$covid != "Non-Covid",], scale = cormat)

p <- ncol(Xa)
q <- ncol(Ya)
na <- nrow(Xa)
nt <- nrow(Xt)
Kmax <- min(p, q, nt-1)
covariates_a <- covariates[rownames(Xa),]
covariates_t <- covariates[rownames(Xt),]


#### 3. s3CA  ####

res_s3CA <- cCCA(Xa = Xa, Ya = Ya, Xt = Xt, Yt = Yt, eta  = eta_assign,
                sparse_ctst = seq(0.2, 1, len=20), sparse_trt = seq(0.2, 1, len=40), sparse_ctrl = seq(0.2, 1, len=40))
CCA_ctst <- res_s3CA$CCA_ctst
CCA_ctrl <- res_s3CA$CCA_ctrl
CCA_trt <- res_s3CA$CCA_trt

K_single <- K_t <- K_a <- 1
cCCX <- Xt%*%as.matrix(CCA_ctst$u[,K_single])
cCCY <- Yt%*%as.matrix(CCA_ctst$v[,K_single])

df_score <- data.frame(cCCX = cCCX, cCCY = cCCY, covariates_t)

df_score_t <- data.frame(CCtX = Xt%*%as.matrix(CCA_trt$u[,K_t]), 
                         CCtY = Yt%*%as.matrix(CCA_trt$v[,K_t]), covariates_t)

df_score_a <- data.frame(CCcX = Xa%*%as.matrix(CCA_ctrl$u[,K_a]), 
                         CCcY = Ya%*%as.matrix(CCA_ctrl$v[,K_a]), covariates_a)


#### 4. Visualization ####
# 4.1. scree
p_scree <- ggplot(data = data.frame(Component = 1:Kmax, Variation = CCA_ctst$d/sum(CCA_ctst$d)))+
  geom_point(aes(x = Component, y = Variation))+
  theme_bw()+ 
  labs(y = "Proportion of explained variation")+
  theme(
        panel.grid = element_blank(),
        text = element_text(size = 14),  # Adjust text size
        axis.title = element_text(size = 16),  # Adjust axis title size
        axis.text = element_text(size = 12)
      )
print(p_scree)
# ggsave("../inst/SavedFigures/COVID_scree.pdf", plot = p_scree, width = 5, height = 4)


# 4.2. covariance discrepancy
K <- 30
Wa <- Xa%*%CCA_ctst$u[,1:K]
Za <- Ya%*%CCA_ctst$v[,1:K]
Wt <- Xt%*%CCA_ctst$u[,1:K]
Zt <- Yt%*%CCA_ctst$v[,1:K]

cor_disc <- data.frame(Component = rep(1:K, 2),
                       Cov = c(diag(cov(Wa, Za)), diag(cov(Wt, Zt))),
                       Type = rep(c("Control", "Treatment"), each = K))
p_disc <- ggplot(data = cor_disc) +
  geom_line(aes(x = Component, y = Cov, color = Type))+
  labs(y = "Cov(Xu, Yv)")+
  theme_bw()+
  theme(
        panel.grid = element_blank(),
        text = element_text(size = 14),  # Adjust text size
        axis.title = element_text(size = 16),  # Adjust axis title size
        axis.text = element_text(size = 12),
        legend.position = "bottom"
      )+
  scale_color_aaas()
# ggsave("../inst/SavedFigures/COVID_disc.pdf", plot = p_disc, width = 6, height = 5)
print(p_disc)


# 4.3. score scatter nuisance removal
nuisance <-  "icu"
ARI_removal <- c(spec_clust(Ya%*%CCA_ctrl$v[,1], group = df_score_a[,nuisance]),
                 spec_clust(Yt%*%CCA_trt$v[,1], group = df_score_t[,nuisance]),
                 spec_clust(Yt%*%CCA_ctst$v[,1], group = df_score[,nuisance]))
ARI_removal <- paste("ARI(ICU) = ", ARI_removal)
names(ARI_removal) <- c("Control","Treatment","Contrastive")

# 4.4. score scatter continuous interested regression
interested_y <- "Platelet_K.uL"
pval_sig <- c(summary(score_reg(x = data.frame(x = df_score[,"cCCY"]),
                                y = df_score[,interested_y]))$coefficients[2,"Pr(>|t|)"],
              summary(score_reg(x = data.frame(x = df_score_t[,"CCtY"]),
                                y = df_score_t[,interested_y]))$coefficients[2,"Pr(>|t|)"],
              summary(score_reg(data.frame(x = df_score_a[,"CCcY"]), 
                                y = df_score_a[,interested_y]))$coefficients[2,"Pr(>|t|)"])
pval_sig <- round(pval_sig, 4)
pval_sig <- paste("P-value = ", pval_sig, sep = "")
names(pval_sig) <- c("Contrastive","Treatment","Control")

p_reg <- ggplot(data = data.frame(df_score[,c("cCCY","icu",interested_y)]),
                aes_string(x = "cCCY",y = interested_y))+
  geom_point(aes_string(color = "icu" ))+ 
  geom_smooth(method = "lm", se = TRUE)+
  geom_text(data = data.frame(a=c(3,3),b=c(700, 600),
                              label=c(ARI_removal["Contrastive"], pval_sig["Contrastive"])), 
            aes(x = a, y = b, label = label))+
  theme_bw()+
  labs(title = "Contrastive", x = "Yv", color = "")+
  theme(panel.grid = element_blank(),
        text = element_text(size = 14),  # Adjust text size
        axis.title = element_text(size = 16),  # Adjust axis title size
        axis.text = element_text(size = 12),
        legend.position = "right")+
  scale_color_d3()
p_reg_t <- ggplot(data = data.frame(df_score_t[, c("CCtY", "icu",interested_y)]),
                aes_string(x = "CCtY",y = interested_y))+
  geom_point(aes_string(color = "icu" ))+  
  geom_smooth(method = "lm", se = TRUE)+
  geom_text(data = data.frame(a=c(-10,-10),b=c(700, 600),
                              label=c(ARI_removal["Treatment"], pval_sig["Treatment"])), 
            aes(x = a, y = b, label = label))+
  theme_bw()+
  labs(title = "COVID", x = "Yv", color = "")+
  theme(panel.grid = element_blank(),
        text = element_text(size = 14),  # Adjust text size
        axis.title = element_text(size = 16),  # Adjust axis title size
        axis.text = element_text(size = 12),
        legend.position = "right")+
  scale_color_d3()
p_reg_a <- ggplot(data = data.frame(df_score_a[, c("CCcY", "icu",interested_y)]),
                aes_string(x = "CCcY",y = interested_y))+
  geom_point(aes_string(color = "icu" ))+ 
  geom_smooth(method = "lm", se = TRUE)+
  geom_text(data = data.frame(a=c(-15,-15),b=c(313, 267),
                              label=c(ARI_removal["Control"], pval_sig["Control"])), 
            aes(x = a, y = b, label = label))+
  theme_bw()+
  labs(title = "Non COVID", x = "Yv", color = "")+
  theme(panel.grid = element_blank(),
        text = element_text(size = 14),  # Adjust text size
        axis.title = element_text(size = 16),  # Adjust axis title size
        axis.text = element_text(size = 12),
        legend.position = "right")+
  scale_color_d3()

p_score <- ggarrange(p_reg_t, p_reg_a, p_reg, common.legend = T, 
                     legend = "bottom", nrow=1, align = "hv")
print(p_score)
# ggsave("../inst/SavedFigures/COVID_score.pdf", plot = p_score, width = 12, height = 6)
           

# 4.5 loading plot
p_loading <- ggplot(data = data.frame(index = 1:ncol(Yt),
                                      Loadings = sort(abs(CCA_ctst$v[,1])),
                                      mark = rep(c("Others","Top 11"), c(ncol(Yt)-11,11))))+
  geom_point(aes(x=index, y=Loadings, color = mark))+
  theme_bw()+
  labs( x = "Proteins", y = "Absolute values of loadings", color = "")+
  theme(panel.grid = element_blank(),
        text = element_text(size = 14),  # Adjust text size
        axis.title = element_text(size = 16),  # Adjust axis title size
        axis.text = element_text(size = 12),
        legend.position = "bottom")+
  scale_color_d3()
print(p_loading)

# ggsave("../inst/SavedFigures/COVID_loadings.pdf", plot = p_loading, width = 5, height = 4)

print(colnames(Yt)[order(abs(CCA_ctst$v[,1]), decreasing = T)][1:11])
```

